Computing the bicharacter of huge subspaces of polynomials

### 2019-05-06 article/article.tex
Add content in context section

	\section{Introduction}
	This package was developped at the occasion of a collaboration with François Bergeron to study the character of the space $\harmonics_{k,n}$ of diagonal harmonic polynomials on $k$ sets of $n$ variables, as a graded bi $\GLr$-$\Sn$  module. The main achievement is a calculation of this character for $n=6$ and $r=5$ (and therefore any $r$) in 45 minutes (and ~15Gb), which led to conjectures on its general value.

@@

	Consider the polynomial ring $P=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables: \pauline{small $x$ for variables, capital $X$ for sets of variables? }
	$$(X_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $X_i$ the i-th set of variable.

	The polynomial ring $P$ is endowed with an action of the symmetric group $\Sn$ by diagonal permutation of the columns, and $\GLr$ by diagonal action on the rows. The invariant ring $P^{\Sn}$ is also known as the ring of multisymmetric functions. The co-invariant ring $P_{\Sn}$ is the quotient of $R$ by the ideal $\langle {P^{\Sn}}^+ \rangle$ \pauline{+ = without constant terms ?}. It inherits from $P$ a structure of graded $\GLr$-$\Sn$ bi-module.

	The grading can be refined into a multigrading by rows. The multigrading is preserved by the action of $\Sn$. The multigrading is not preserved by the action of $\GLr$, but instead carries exactly the information about the $\GLr$ character. It's thus equivalent to consider the $\GLr$-$\Sn$ bicharacter or the $\Sn$ multigraded character of $P_{\Sn}$.

	When $k=1$, the action of $GL_1$ is trivial; $P_{\Sn}$ is of dimension $n!$ and carries the usual graded representation of $\Sn$. Its character is given by a Hall-Littlewood polynomial.
	When $k=2$, $P_{\Sn}$ is of dimension $(n+1)^{n-1}$ ($=16807$ for $n=6$); as a consequence of Haiman's proof of the $n!$-conjecture, the bigraded character is a Macdonald polynomial.

	Throughout this notebook, we implicitly use the Frobenius map, and identify $\Sn$ and $\GLr$ characters with their associated symmetric functions. \\

@@

	Note that $V_n:=$ $\prod_{i<j}(x_i-x_j)$ is harmonic and that if $f$ is harmonic then $\partial_i f$ is also harmonic. \\

	Let now go back to $X$ containing an arbitrary number $r$ of sets of variables.
	We define the \emph{differential operators} $(D_\alpha)_{\alpha\in \NN^r}$:
	$$D_{(\alpha_1, \dots, \alpha_r)} := \partial_{x_{11}}^{\alpha_1}\partial_{x_{r1}}^{\alpha_r} + \dots + \partial_{x_{1n}}^{\alpha_1}\partial_{x_{rn}}^{\alpha_r}$$

	For example, $r=3$ and we denote $x, y$ and $z$ the three sets of variables of $X$, then $D_{(3,0,4)} := \partial_{x_1}^3\partial_{z_1}^4 + \dots + \partial_{x_n}^3\partial_{z_n}^4$.

	The space of the diagonal harmonic polynomials $\harmonics_{n,r}$ is the joint kernel of the $(D_\alpha)_\alpha$. And $\harmonics_{n,r}$ contains $\harmonics_{n,1} = \harmonics_{n}$.

@@

	It is well known (\cite{Haiman2002} and \cite{Bergeron2009}) that the harmonics $\harmonics_{n,1}$ are generated by the Vandermonde determinant $V_n$ and closed by partial derivatives. Moreover, the diagonal harmonics $\harmonics_{n,r}$ are generated by $\harmonics_{n,1}$ and closed by the polarization operators.

	\begin{definition} \pauline{how to introduce them ?}
		The \emph{Steenrod operator} $St_{x_a}^k$ for $k>1$ is given by
		$$St_{x_a}^k = \sum_{i=1}^{n} x_{ai} \partial_{x_{ai}}^k$$
		where $x_a$ is a set of variables of $X$. 
	\end{definition}

>>>>

	\section{Introduction}
	This package was developped at the occasion of a collaboration with François Bergeron to study the character of the space $\harmonics_{k,n}$ of diagonal harmonic polynomials on $k$ sets of $n$ variables, as a graded bi-$\GLr \times \Sn$-module. The main achievement is a calculation of this character for $n=6$ \pauline{+ some cases for $n=7$ ?} and $r=5$ (and therefore any $r$) in 45 minutes (and ~15Gb), which led to conjectures on its general value.

@@

	Consider the polynomial ring $P=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables: \pauline{small $x$ for variables, capital $X$ for sets of variables? } \pauline{introduce inert variables now or later?}
	$$(X_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $X_i$ the i-th set of variables.

	The polynomial ring $P$ is endowed with an action of the symmetric group $\Sn$ by diagonal permutation of the columns and of $\GLr$ by diagonal action on the rows. The invariant ring $P^{\Sn}$ is also known as the ring of multisymmetric functions. The co-invariant ring $P_{\Sn}$ is the quotient of $P$ by the ideal $\langle {P^{\Sn^+}} \rangle$ of the multisymmetric functions without constant term. It inherits from $P$ a structure of graded $\GLr \times \Sn$ bi-module.

	The grading can be refined into a multigrading by rows. The multigrading is preserved by the action of $\Sn$. The multigrading is not preserved by the action of $\GLr$, but instead carries exactly the information about the $\GLr$ character. It's thus equivalent to consider the $\GLr \times \Sn$ bicharacter or the $\Sn$ multigraded character of $P_{\Sn}$.

	When $k=1$, the action of $GL_1$ is trivial; $P_{\Sn}$ is of dimension $n!$ and carries the usual graded representation of $\Sn$. Its character is given by a Hall-Littlewood polynomial.
	When $k=2$, $P_{\Sn}$ is of dimension $(n+1)^{n-1}$ ($=16807$ for $n=6$); as a consequence of Haiman's proof of the $n!$-conjecture (see \cite{Haiman2002}), the bigraded character is a Macdonald polynomial.

	Throughout this notebook, we implicitly use the Frobenius map, and identify $\Sn$ and $\GLr$ irreducible characters with their associated Schur symmetric functions. \\

@@

	Note that the Vandermonde determinant $V_n:=$ $\prod_{i<j}(x_i-x_j)$ is harmonic. It is also easy to see that if $f$ is harmonic then $\partial_{x_i} f$ is also harmonic for all $x_i \in X$. \\

	Going back to an arbitrary number $r$ of sets of variables, we define the \emph{differential operators} $(D_\alpha)_{\alpha\in \NN^r}$:
	$$D_{(\alpha_1, \dots, \alpha_r)} := \partial_{x_{11}}^{\alpha_1}\dots\partial_{x_{r1}}^{\alpha_r} + \dots + \partial_{x_{1n}}^{\alpha_1}\dots\partial_{x_{rn}}^{\alpha_r} = \sum_{j=1}^{n} \prod_{i=1}^{r} \partial_{x_{ij}}^{\alpha_i}$$

	For example, take $r=3$ and denote $x, y$ and $z$ the three sets of variables of $X$, then $D_{(3,0,4)} := \partial_{x_1}^3\partial_{z_1}^4 + \dots + \partial_{x_n}^3\partial_{z_n}^4$.

	The space of the diagonal harmonic polynomials $\harmonics_{n,r}$ is the joint kernel of the $(D_\alpha)_\alpha$. Note that $\harmonics_{n,r}$ contains $\harmonics_{n,1} = \harmonics_{n}$.\\

	The previous definition means that the diagonal harmonic polynomials are solutions of differential equations systems. They can also be obtained by construction using polarization operators.

@@

	It is well known (\cite{Haiman2002} and \cite{Bergeron2009}) that the harmonics $\harmonics_{n,1}$ are generated by the Vandermonde determinant $V_n$ and closed by partial derivatives. Moreover, the diagonal harmonics $\harmonics_{n,r}$ are generated by $\harmonics_{n,1}$ and closed by the polarization operators. \\

	The diagonal harmonics $\harmonics_{n,r}$ have been studied a lot until now. The character in the case $r=2$ is due to Garsia and Haiman (see \cite{GarsiaHaiman1993}) and some progress in the case $r=3$ had recently been made by Bergeron and Préville-Ratelle (see \cite{BergeronPreville2012}), but the general case is still unknown. Therefore, the computation of those characters should lead to a better understanding of those spaces.

	In their work, Garsia and Haiman also introduced a generalized version of the Vandermonde determinant and a generalization of the diagonal harmonics for two sets of variables called the Gasia-Haiman modules. \\

	F. Bergeron recently introduced another generalization of the diagonal harmonics, with inert variables. We add a new row $\theta = (\theta_1, \theta_2, \dots, \theta_n)$ of variables in $X$. The symmetric group still acts diagonally on the rows of $X$, but $GL_r$ now acts only on the rows containing non-inert variables. Like before, we consider the smallest subspace containing a certain determinant and closed by partial derivation and polarization, except that this time the inert variables do not intervene in the operators. The determinant used in this case is defined as follow.

	Let $\lambda$ be a partition of $n$, we also denote $\lambda$ the Ferrer's diagram associated to the partition $\lambda$ and each cell of $\lambda$ is given by the coordinates of its bottom left corner. Then the determinant $V_{\lambda}$ associated to $\lambda$ is given by $$V_{\lambda} = \det (x_i^a \theta_i^b)_{1 \leq i \leq n \atop (a,b) \in \lambda} .$$
	For example, the partition $21$ contains the cells $(0,0)$ and $(1,0)$ in the first row, and the cell $(0,1)$ in the second row, and $V_{21} = -x_2\theta_1 + x_3\theta_1 + x_1\theta_2 - x_3\theta_2 - x_1\theta_3 + x_2\theta_3$. \\

	Since those subspaces are completely new, computing some characters on Sage helps understanding them better and allows to find some properties of thoses spaces. 
	Plus, experimentation helped finding the right definition we needed for this generalization.

	\begin{definition} \pauline{find the right place}
		The \emph{Steenrod operator} $St_{x_a}^k$ for $k>1$ is given by
		$$St_{x_a}^k = \sum_{i=1}^{n} x_{ai} \partial_{x_{ai}}^k$$
		where $x_a$ is a set of variables of $X$. 
	\end{definition}
